---
title: "Mapping Australia’s Coastal Defences: Understanding the Gaps and the Need for a National Dataset"
author: "NESP Climate Systems Hub researchers and knowledge brokers [info@nesp2climate.com.au](mailto:info@nesp2climate.com.au)"
date: "`r Sys.Date()`"
output:
  html_document: default
  word_document: default
---

```{css, echo=FALSE}
pre code {
  display: block;
  max-height: 6em;   /* Approximately 3 lines */
  overflow-y: scroll;  /* Scroll vertically if content exceeds height */
  white-space: pre-wrap; /* Prevent horizontal scrolling */
  border: 1px solid #ddd;
  padding: 10px;
  background-color: #f9f9f9;
}```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=10, fig.height=7)
```

Australia's coastline is facing increasing pressures from erosion, exacerbated by rising sea levels and extreme weather events. To determine which parts of the coast are vulnerable, it’s crucial to identify areas protected by coastal structures such as sea walls, groynes, and wharfs.

[The Climate Systems Hub, through its knowledge brokers, is helping to address this issue by enquiring with state governments about available spatial data on coastal protection structures](http://hdl.handle.net/102.100.100/660109?index=1). Ideally, these datasets would include vital information such as the location, height, and type of each structure.

Currently, Victoria, Tasmania, Queensland, and Western Australia provide standardised publicly accessible datasets on their coastal protection structures. However, this data is not readily available for South Australia, the Northern Territory, or New South and there is insufficient information regarding potential updates to these datasets. A comprehensive national dataset would help researchers and decision makers assess the risks to Australia’s coastlines and guide efforts to protect vulnerable regions.

A recent report from the [Insurance Council of Australia, titled Actions of the Sea](https://insurancecouncil.com.au/resource/actions-of-the-sea-vulnerability-study/#:~:text=The%20report%20acknowledges%20the%20growing,climate%20increases%20exposure%20and%20risk.&text=Climate%20change%20is%20driving%20rising,inundation%2C%20erosion%2C%20and%20recession.), highlights the need for consistency in how coastal defence data is gathered and shared. It recommends that state and local governments develop spatial databases of coastal defences according to a standard framework.  
The framework should capture key details, such as:  
  
- Location and type of coastal protection.  
- Studies evaluating the function and hazard mitigation performance of the structures.  
- Key design parameters, including storm return periods and allowances for future sea level rise.  
- Construction cost and design life information.  

The state-based datasets on coastal structures vary in detail and naming conventions, and a limitation is that structures are built yearly, yet these datasets aren’t regularly updated. Depending on the quality and level of detail, the data can inform applications at different scales—from first pass scientific research at the national scale to detailed first-pass analysis to inform planning policy decisions.

Existing national datasets, such as Smartline, are being used to help fill current gaps in our datasets of coastal defences. Smartline is a national product that identifies whether a structure exists or not. OpenStreetMap (OSM), a collaborative project where users contribute geographic information, also includes mapping of coastal structures in Australia. Although OSM is not yet comprehensive, it serves as a useful starting point for understanding where protections exist—and where they do not. By exploring OSM contributions, we aim to inform efforts to collect standardised data and make it more accessible to coastal researchers, engineers, and policy managers.

By collating and highlighting gaps in coastal defence data, the analysis presented complements NESP’s outreach efforts. Together, these initiatives will strengthen our understanding of coastal protection and help identify the areas most vulnerable to erosion, guiding future efforts to safeguard Australia’s shores.

Below, we present `R` code analysis of the publicly available data. If you scroll to the end, you'll find a interactive map comparing the national OSM and state based data [OR GO TO THE MAP NOW](#target-map), along with a **Call to Action: Get Involved!** to support our efforts to assemble a national dataset of coastal structures.

To download the processed national merged data file generated in the code below, click the following button:

<a href="data/National_Coastal_structures_20241024.gpkg" download>
  <button>Download GPKG geospatial file</button>
</a>

The file can be read into GIS software.

Please Cite as **O'Grady, Julian; Trenham, Claire; Morris, Rebecca; & NESP Climate Systems Hub Knowledge Brokers (2024): Mapping Australia's Coastal Defences Structures: Understanding the Gaps and the Need for a National Dataset. v2. CSIRO. Service Collection.** [http://hdl.handle.net/102.100.100/660109?index=1](http://hdl.handle.net/102.100.100/660109?index=1)

In this research-quality dataset, a uniform naming convention is presented, lines are converted to polygons (with a width of 1m), and the data source is noted. There are important limitations in the data which are presented in the analysis, including no government-supplied data for NSW, NT, or SA, no sea walls in the TAS dataset, and the OSM data is provided by citizen scientists without extensive professional review of the labelled coastal structure type.

All information and data were generated for research purposes. While every effort has been made to ensure data accuracy, the developers make no warranty (expressed or implied) regarding the accuracy, completeness, function, or capability of the code and associated data. The data is not provided as professional advice, and the user assumes any risk associated with its use, possession, or reliance on the content for decision-making. The developers expressly disclaim liability for any loss arising from the use of this code and the information contained therein. The code incorporated here is general in nature, and more detailed studies may be required for individual regions.

### Identified coastal structure datasets

The NESP knowledge brokers were able to identify the datasets in the following table. The data was downloaded to and placed in the `/data` folder in this repository along with readme files that explain the data download. In the R scripts below, the data is read in, interrogated, and a merged product is provided to inform first-pass assessments of national coastal structures.

| **Dataset** | **Public dataset links** |
|------------|-------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| VIC        | [VIC Data](https://discover.data.vic.gov.au/dataset/coastal-protection-structures) download |
| TAS        | [TAS Data](https://www.thelist.tas.gov.au/app/content/data/geo-meta-data-record?detailRecordUID=195c6de2-53e8-4792-84fa-5ab1590b2f8c) dowloaded and filtered for marine features |
| QLD        | Searched for "Marine infrastructure lines" on qldspatial, [QLD Spatial](https://qldspatial.information.qld.gov.au/catalogue/custom/search.page) |
| WA         | [WA Data](https://catalogue.data.wa.gov.au/dataset/coastal-infrastructure-lines) Downloaded/viewed/exported via QGIS WFS [WA WFS](https://public-services.slip.wa.gov.au/public/services/SLIP_Public_Services/Infrastructure_and_Utilities_WFS/MapServer/WFSServer) and exported as geojsons |
| OSM        | [OpenStreetMap](https://www.openstreetmap.org/), Downloaded via Overpass Turbo Query: [Overpass Query](https://overpass-turbo.eu/?template=key-value&key=man_made&value=groyne#) |
| Smartline  | [Smartline](https://coastadapt.com.au/coastadapt-interactive-map), Sharples, C., & Mount, R. E. (2009). The Australian coastal Smartline geomorphic and stability map version 1: manual and data dictionary. |


### Analysis of coastal structures 

The sections of code below provide a working example of processes for geospatial data on coastal structures sourced from the table above.

From OpenStreetMap (OSM), the key-value pairs used to filter relevant OSM data include:
  
1.&nbsp;&nbsp;&nbsp;&nbsp;**man_made = groyne:** A rigid structure built from the shore to interrupt water flow and limit sediment movement, often used to prevent beach erosion. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Dgroyne  
2. &nbsp;&nbsp;&nbsp;&nbsp;**man_made = embankment:** An artificial steep slope built along coastlines or rivers to prevent water from overflowing and flooding adjacent land. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Dembankment  
3. &nbsp;&nbsp;&nbsp;&nbsp;**man_made = dyke:** A raised structure designed to restrict water flow, typically built to protect coastal land from flooding. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Ddyke  
4. &nbsp;&nbsp;&nbsp;&nbsp;**man_made = quay:** A solid platform along the shore or a riverbank, built for loading and unloading ships. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Dquay  
5. &nbsp;&nbsp;&nbsp;&nbsp;**man_made = breakwater:** A structure built offshore to protect a coast or harbour from the force of waves. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Dbreakwater  
6. &nbsp;&nbsp;&nbsp;&nbsp;**man_made = pier:** A raised platform extending into the sea or a body of water, commonly used for docking boats or recreational purposes. https://wiki.openstreetmap.org/wiki/Tag:man_made%3Dpier  
7. &nbsp;&nbsp;&nbsp;&nbsp;**bridge = boardwalk:** A wooden or raised path usually found along coastal areas, providing pedestrian access over marshes or water. https://wiki.openstreetmap.org/wiki/Tag:bridge%3Dboardwalk  
8. &nbsp;&nbsp;&nbsp;&nbsp;**barrier = retaining_wall:** A wall constructed to prevent soil or coastal land from eroding or sliding into the water. https://wiki.openstreetmap.org/wiki/Tag:barrier%3Dretaining_wall  
9. &nbsp;&nbsp;&nbsp;&nbsp;**barrier = wall:** A generic wall structure that can also be used along coasts to define property or act as flood protection. https://wiki.openstreetmap.org/wiki/Tag:barrier%3Dwall  
10. &nbsp;&nbsp;&nbsp;&nbsp;**wall = flood_wall:** A specific type of wall designed to keep coastal or river waters from flooding inland areas. https://wiki.openstreetmap.org/wiki/Tag:wall%3Dflood_wall  
11. &nbsp;&nbsp;&nbsp;&nbsp;**wall = seawall:** A structure built parallel to the coast to protect the shore from erosion and storm surges. https://wiki.openstreetmap.org/wiki/Tag:wall%3Dseawall  
12. &nbsp;&nbsp;&nbsp;&nbsp;**emergency = lifeguard:** A designated spot or station for lifeguards, typically located on beaches to ensure water safety. https://wiki.openstreetmap.org/wiki/Tag:emergency%3Dlifeguard  
13. &nbsp;&nbsp;&nbsp;&nbsp;**leisure = slipway:** A ramp leading into the water, used for launching or retrieving boats. https://wiki.openstreetmap.org/wiki/Tag:leisure%3Dslipway  


For each of the key-value pairs, the OSM Overpass API was used to query and extract GeoJSON files for these 13 key-value pairs:
- Overpass Turbo Query: https://overpass-turbo.eu/?template=key-value&key=man_made&value=groyne#

The script performs the following steps:  
  
1. Load and crop the coastline vector data for Australia.  
2. Buffer the coastline by 200 meters to account for nearby coastal structures.  
3. List and read relevant OpenStreetMap (OSM) GeoJSON files containing coastal structures, downloaded from the Overpass API above.  
4. Extract polygon and line geometries from the OSM files.  
5. Clean and convert line geometries to polygons with a width of 1 meter.  
6. Intersect coastal structures with the Australian coastline.  
7. Assign unique Feature IDs (FIDs) to the intersected features.  
8. Combine and filter valid geometries into a unified dataset.  
9. Write the final spatial vector object to a GeoPackage file.  
10. Calculate area statistics (mean, min, max, and standard deviation) for coastal structures.  
11. Load and compare Victoria’s coastal structures dataset with the OSM data.  
12. Calculate summary statistics for Victorian data, grouped by structure type.

The output is a GeoPackage file containing the filtered and processed coastal structures along the Australian coastline.
This file can be used for further analysis or mapping of coastal defences in Australia.

```{r env}
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(terra))
suppressPackageStartupMessages(library(sf))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))

datadir = "data/"
```
#### The coast

Coastal structures are defined here as structures being within 100 m from the shoreline. The coast was made via adding a 200m buffer around a 100m OSM coastline of Australia from the code in https://doi.org/10.25919/mapj-bq95 

```{r aus coast}
aus_coast = vect(paste0(datadir,"OSM_AUScoastBuffer200m",".gpkg"))
```

Here we read in the OSM geoJSONs download on 29th Aug 2024.
```{r input osm files}
OSM_geofiles = list.files(paste0(datadir,"OSM_shapes_20240829"),full.names = TRUE)
OSM_geofiles = paste0(datadir,"OSM_shapes_20240829/export (",1:13,").geojson")

filetxt <- c(
  "man_made:groyne",
  "man_made:embankment",
  "man_made:dyke",
  "man_made:quay",
  "man_made:breakwater",
  "man_made:pier",
  "bridge:boardwalk",
  "barrier:retaining_wall",
  "barrier:wall",
  "wall:flood_wall",
  "wall:seawall",
  "emergency:lifeguard",
  "leisure:slipway"
)

#open street map coastal geojsons

geoms = lapply(OSM_geofiles,function(x) sapply(svc(x),function(x) geomtype(x)))

n = length(OSM_geofiles)

#some shapes are lines, here is a list of lines
OSM_cps_pl = lapply(1:n,function(x) {if(any(geoms[[x]]=="polygons")) {y=NULL;y = svc(OSM_geofiles[x])[[which(geoms[[x]]=="polygons")]];y$filetype = filetxt[x];y$drawn = "polygons";y}})
#some shapes are polygons, here is a list of polygons
OSM_cps_ll = lapply(1:n,function(x) {if(any(geoms[[x]]=="lines"))    {y=NULL;y = svc(OSM_geofiles[x])[[which(geoms[[x]]=="lines")]];y$filetype = filetxt[x];y$drawn = "lines";y}})

```

Clean the lists and convert lines to polygons a with width of 1m.

```{r tidy osm shapes}
OSM_cps_p = vect(OSM_cps_pl[which(!sapply(OSM_cps_pl,is.null))])
touch = relate(OSM_cps_p,aus_coast,"intersects",pairs = TRUE)
OSM_cps_p = OSM_cps_p[unique(touch[,1])]
  
OSM_cps_l = vect(OSM_cps_ll[which(!sapply(OSM_cps_ll,is.null))])
touch = relate(OSM_cps_l,aus_coast,"intersects",pairs = TRUE)
OSM_cps_l = buffer(OSM_cps_l[unique(touch[,1])],0.5)


```

Bring the datasets together and keep only the shapes within 200km of the OSM coast.
```{r bring together shapes}
OSM_cps_coast = rbind(OSM_cps_l,OSM_cps_p)
OSM_cps_coast = buffer(OSM_cps_coast,0)

#OSM_cps_coast = OSM_cps_coast[which(is.valid(OSM_cps_coast))]
#cols_to_keep <- as.numeric(which(!sapply(OSM_cps_coast, function(x) all(is.na(x) | x == ""))))
#OSM_cps_coast_out <- OSM_cps_coast[,cols_to_keep]
#OSM_cps_coast_out <- OSM_cps_coast_out[,-2]
OSM_cps_coast_out = OSM_cps_coast

values(OSM_cps_coast_out) = NA
OSM_cps_coast_out$filetype = OSM_cps_coast$filetype
OSM_cps_coast_out$FID = 1:length(OSM_cps_coast_out)

OSM_cps_coast_out$MergeName = OSM_cps_coast$filetype
writeVector(OSM_cps_coast_out,paste0(datadir,"OSM_coastalStructures.geojson"), filetype = "GeoJSON",overwrite=TRUE)

```

Because some shapes are lines and some polygons, lines were converted to a 1m wide polygon

```{r}

calculate_area_summary <- function(vect_data, group_by_col) {
  # Calculate area in square meters
  vect_data$area_m <- expanse(vect_data, unit = "m")
  
  # Group by the specified column and summarise
  summary_stats <- values(vect_data) %>%
    group_by(.data[[group_by_col]]) %>%
    summarise(
      count = n(),
      mean_area_m = round(mean(area_m), 2),
      min_area_m = round(min(area_m), 2),
      max_area_m = round(max(area_m), 2),
      std_area_m = round(sd(area_m), 2)
    )
  
  # Rename the grouping column for clarity
  names(summary_stats)[1] <- group_by_col
  
  # Display the summary statistics
  return(knitr::kable(summary_stats))
}

calculate_area_summary(OSM_cps_coast_out,"filetype")

```

```{r}
smartline_data = buffer(vect("data/smartline_basic_artificial_shores"),0.5)
smartline_data$MergeName = smartline_data$erode_v
```


#### Comparison to the state government data. 
The Victoria Governments has published the most complete public database of coastal structures https://discover.data.vic.gov.au/dataset/coastal-protection-structures. Here we compare the OSM shapes to the victorian Database.

The Victorian datasets has some similar (Groyne and Breakwater), and some different (Seawall), label types for features compared to OSM. 

```{r}
vicdata <- project(buffer(vect("data/victoria/COASTS"),0.5),OSM_cps_coast_out)
vicdata$MergeName = vicdata$STRUC_TYPE

#calculate_area_summary(crop(OSM_cps_coast_out,ext(vicdata)),"filetype")
calculate_area_summary(vicdata,"STRUC_TYPE")

```


The Tasmanian data was sourced from https://www.thelist.tas.gov.au/app/content/data/geo-meta-data-record?detailRecordUID=195c6de2-53e8-4792-84fa-5ab1590b2f8c

Here is a summary of the data for Tasmania

Note, there are no records of sea wall in the Tasmanian dataset, E.g. the extensive rock revietment seawall in [Burnie](https://www.google.com/maps/@-41.0619778,145.9143705,3a,60y,38.07h,75.81t/data=!3m7!1e1!3m5!1sBm08_qwTv0wzbqu1AuQ6Kg!2e0!6shttps:%2F%2Fstreetviewpixels-pa.googleapis.com%2Fv1%2Fthumbnail%3Fcb_client%3Dmaps_sv.tactile%26w%3D900%26h%3D600%26pitch%3D14.190940796250743%26panoid%3DBm08_qwTv0wzbqu1AuQ6Kg%26yaw%3D38.07064762868976!7i16384!8i8192?coh=205410&entry=ttu&g_ep=EgoyMDI0MTAyMS4xIKXMDSoASAFQAw%3D%3D)

```{r}
tas_line = vect(paste0(datadir,"tasmania/built_infrastructure_line_statewide.shp"))
tas_poly = vect(paste0(datadir,"tasmania/built_infrastructure_poly_statewide.shp"))

tasdata = project(rbind(buffer(tas_line[tas_line$INFTY1_TXT == "Marine"],0.5),tas_poly[tas_poly$INFTY1_TXT == "Marine"]),OSM_cps_coast_out)
tasdata$MergeName = tasdata$INFTY2_TXT

calculate_area_summary(tasdata,"INFTY2_TXT")

```

The Queensland data was obtained by searching for "Marine infrastructure lines" on qldspatial https://qldspatial.information.qld.gov.au/catalogue/custom/search.page
The files where then downloaded from an email sent by an automated API. 

```{r}
qld_lines = vect(paste0(datadir,"queensland/Marine_infrastructure_lines"))
qlddata = buffer(project(qld_lines,rast()), 0.5)

qlddata$MergeName = qlddata$feature_ty

calculate_area_summary(qlddata,"feature_ty")

```

The Western Australian data was downloaded/viewed/exported via QGIS WFS https://public-services.slip.wa.gov.au/public/services/SLIP_Public_Services/Infrastructure_and_Utilities_WFS/MapServer/WFSServer and exported as geojsons

```{r}
wa_lines = vect(paste0(datadir,"western_australia/Coastal_Infrastructure_polygons_DOT-020.geojson"))
wa_lines$AssumStuct = NA
wa_lines$AssumStuct[wa_lines$structype == "BKW"] = "Breakwater"
wa_lines$AssumStuct[wa_lines$structype == "GRN"] = "Groyne"
wa_lines$AssumStuct[wa_lines$structype == "JET"] = "Jetty"
wa_lines$AssumStuct[wa_lines$structype == "SWL"] = "Seawall"
wa_lines$AssumStuct[wa_lines$structype == "WHF"] = "Wharf"
wa_lines$AssumStuct[wa_lines$structype == "MS_Misc"] = "Other"
wa_lines$AssumStuct[wa_lines$structype == "BDW"] = "Boardwalk"
wa_lines$AssumStuct[wa_lines$structype == "OBW"] = "Offshore breakwater"
wa_lines$AssumStuct[wa_lines$structype == "Buoy"] = "Buoy"
wa_lines$AssumStuct[wa_lines$structype == "BLR"] = "Boat lifter"
wa_lines$AssumStuct[wa_lines$structype == "CSW"] = "Causeway"
wa_lines$AssumStuct[wa_lines$structype == "OFP"] = "Outfall Pipe"
wa_lines$AssumStuct[wa_lines$structype == "CHN"] = "Channel"

wadata = wa_lines[!is.na(wa_lines$AssumStuct)]

wadata$MergeName = wadata$structype

calculate_area_summary(wadata,"structype")

```

# Map of the coastal strucutres {#target-map}

Here we map the Victorian, Tasmanian and OSM shapes starting around Geelong in Western Port Phillip Bay. But you are free to zoom around the state or Tasmania, or country and click on a structure to find more info.

```{r leaflet}

# Convert to 'sf' objects for leaflet compatibility
OSM_cps_sf <- st_as_sf(OSM_cps_coast_out)
vicdata_sf <- st_as_sf(vicdata)
tasdata_sf <- st_as_sf(tasdata)  
qlddata_sf <- st_as_sf(qlddata)  
wadata_sf <- st_as_sf(wadata)  

# Define the bounding box for Geelong (approximate coordinates)
geelong_bbox <- st_bbox(c(xmin = 144.30, ymin = -38.25, xmax = 144.45, ymax = -38.10), crs = st_crs(4326))

# Create the leaflet map
leaflet() %>%
  addProviderTiles("OpenStreetMap", group = "OpenStreetMap") %>%
  addProviderTiles("Esri.WorldImagery", group = "Satellite") %>%
  addPolygons(data = OSM_cps_sf, color = "red", weight = 2, opacity = 0.7, fillOpacity = 0.3,
              popup = ~paste("OSM Structure Type:", filetype), group = "OSM Structures") %>%
  addPolygons(data = vicdata_sf, color = "blue", weight = 2, opacity = 0.7, fillOpacity = 0.3,
              popup = ~paste("Victorian Structure:", STRUC_TYPE), group = "Victorian Structures") %>%
  addPolygons(data = tasdata_sf, color = "green", weight = 2, opacity = 0.7, fillOpacity = 0.3,
              popup = ~paste("Tasmanian Structure:", INFTY2_TXT), group = "Tasmanian Structures") %>%
  addPolygons(data = qlddata_sf, color = "maroon", weight = 2, opacity = 0.7, fillOpacity = 0.3,
              popup = ~paste("Queensland Structure:", feature_ty), group = "Queensland Structures") %>%
  addPolygons(data = wadata_sf, color = "yellow", weight = 2, opacity = 0.7, fillOpacity = 0.3,
              popup = ~paste("WA Structure:", structype), group = "WA Structures") %>%
  setView(lng = 144.35, lat = -38.15, zoom = 12) %>%
  addLegend("topright", colors = c("red","blue", "green","maroon","yellow"), labels = c("OSM Coastal Structures","Victorian Coastal Structures", "Tasmanian Structures", "Queensland Structures", "WA Structures"), opacity = 1) %>%
  addLayersControl(
    baseGroups = c("OpenStreetMap", "Satellite"),
    overlayGroups = c("OSM Structures", "Victorian Structures", "Tasmanian Structures", "Queensland Structures", "WA Structures"),
    options = layersControlOptions(collapsed = FALSE)
  )
```


### Generation of a merged National dataset

Below is a table detailing the naming of different structures in the different datasets. 
```{r}
comp_table = read.csv("data/compare_feauture_structure_names.csv")
knitr::kable(comp_table)
all_feat_names = c(paste(as.matrix(comp_table[,-6]))[-53],"Wharf_commercial", "Wharf_noncommercial")
all_feat_names = all_feat_names[all_feat_names != ""]


merge_table = read.csv("data/merge_feauture_structure_names.csv")
nm = nrow(merge_table)
merge_list = strsplit(merge_table[,2],",")
feat_use = unlist(merge_list)
feat_other = paste(sapply(all_feat_names, function(x) if(!any(x == feat_use)) x))
feat_other = feat_other[feat_other != "NULL"][-1]
feat_other = c(feat_other,"Dominantly artificial shores (erodibility unclassified but commonly low)") #include smartline as other because it doesn't identify structure type. 
merge_table[8,2] = paste0(feat_other,collapse = ",")
```

The final national product was generated by merging the strcture names in the datasets using the conventions in the following table 

```{r}

knitr::kable(merge_table)

```

```{r}

formatted_shapes = function(v,CoastStruct,MrgFeatures,dataset){
  MrgFeature = strsplit(MrgFeatures,",")[[1]]
  vs = subset(v, v$MergeName %in% MrgFeature)[,"MergeName"]
  vs$CoastStruct = CoastStruct
  vs$dataset = dataset
  return(vs)
}

vicdata_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = vicdata,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "VIC")))
tasdata_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = tasdata,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "TAS")))
qlddata_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = qlddata,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "QLD")))
wadata_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = wadata,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "WA")))
osm_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = OSM_cps_coast_out,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "OSM")))
slb_uniform = vect(lapply(1:8, function(i) formatted_shapes(v = smartline_data,CoastStruct = merge_table$CoastStruct[i],MrgFeatures = merge_table$MrgFeature[i],dataset = "Smartline")))

national_uniform = vect(list(vicdata_uniform,tasdata_uniform,qlddata_uniform,wadata_uniform,osm_uniform,slb_uniform))

writeVector(national_uniform,"data/National_Coastal_structures_20241024.gpkg",overwrite=TRUE)
```

Summaries the merged product.

```{R}

# Function to calculate area for each CoastStruct and dataset
get_area_by_coaststruct_dataset <- function(data) {
  # Calculate area for each feature
  data$area <- terra::expanse(data)
  
  # Convert to data frame for easier manipulation
  df <- as.data.frame(data)
  
  # Group by CoastStruct and dataset and summarize the total area
  df %>%
    group_by(CoastStruct, dataset) %>%
    summarize(total_area = sum(area, na.rm = TRUE)) %>%
    ungroup()
}

# Apply the function to calculate area
area_data <- get_area_by_coaststruct_dataset(national_uniform)

# Pivot the data to create a wide format with datasets as columns
area_table <- area_data %>%
  pivot_wider(names_from = dataset, values_from = total_area)

# Set a large value for scipen to disable scientific notation
options(scipen = 999)

# Print the resulting table
knitr::kable(area_table)

```
Summary of the OSM data in the merged product based on State Coastal Water Areas AMB 2020: the OSM column identifies international structures.

```{r}
cwa = terra::project(vect("data/Coastal_Waters_areas_(AMB2020)"),osm_uniform)
cwa$state = c("VIC","TAS","QLD","NT","WA","SA","NSW")
osm_uniform2 = osm_uniform
re  = terra::relate(cwa,osm_uniform,"intersects")

for(i in 1:7) osm_uniform2$dataset[which(re[i,])] = cwa$state[i]

# Apply the function to calculate area
area_data <- get_area_by_coaststruct_dataset(osm_uniform2)

# Pivot the data to create a wide format with datasets as columns
area_table <- area_data %>%
  pivot_wider(names_from = dataset, values_from = total_area)

# Set a large value for scipen to disable scientific notation
options(scipen = 999)

# Print the resulting table
knitr::kable(area_table)

```

The review of coastal structure GIS spatial features across different Australian states highlights variations in naming conventions. For example, "Boatlaunch" was used to unify "Slipway" and "Boat Ramp", providing a single term to represent locations where vessels are launched. The names have been corrected to avoid abbreviations or overly simplified names, ensuring descriptive terms like "Breakwater" and "Seawall" are consistently used. Furthermore, all terms have been limited to 10 characters, with no abbreviations or three-letter words. This ensures a consistent and clear naming system that can be applied across all datasets for better integration and usability.

The final summary of shapes are. 

### Where to From Here?

In conclusion, researchers and knowledge brokers within NESP are actively exploring the most effective ways to consolidate and centralise coastal protection data. This could involve mapping efforts through platforms like OpenStreetMap (OSM) or the creation of a national data repository. The benefit of OSM lies in its live, regularly updated format, offering real-time insights. However, it also relies heavily on community participation for data accuracy. Striking the right balance between real-time community input and rigorous research quality is crucial to ensuring coastal protection information is both comprehensive and reliable for future decision-making. This consideration is especially important for supporting high-quality research, engineering assessments, and risk management.

## Call to Action: Get Involved!

The success of this initiative depends on the active involvement of researchers, engineers, policymakers, and the broader community. Whether you're a coastal engineer, a scientist, or simply passionate about protecting Australia’s coastline, there are several ways to contribute:

- **Map coastal structures in OpenStreetMap**: If you're familiar with an area and its coastal defences, consider contributing by adding this information to OSM.
- **Contribute your data**: Researchers and local governments are encouraged to share any datasets related to coastal protection structures. This will help build a more complete picture of the national landscape.
- **Provide feedback**: If you notice gaps, inconsistencies, or have suggestions on how to improve the consolidation efforts, your input is invaluable.
- **Join the conversation**: Engage with NESP Climate Systems Hub knowledge brokers [info@nesp2climate.com.au](mailto:info@nesp2climate.com.au) and coastal communities to discuss the best ways to integrate data and insights into meaningful, centralised resources.

Your participation can directly influence the quality and accessibility of coastal protection data, helping to safeguard our shorelines for future generations. Let's work together to ensure Australia’s coastal regions are resilient and well-prepared for the challenges ahead.
    
     
       
         
  .        
  .

